---
title: "Accelerometry"
description: |
  Analysis of accelerometry summary data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(rmarkdown)
library(tibble)
library(dplyr)
library(ggplot2)
library(purrr)
library(trelliscopejs)
library(stringr)
library(RColorBrewer)
library(colorRamps)
library(ggpubr)
options(dplyr.summarise.inform = FALSE)
```

# Preprocess  

The original accelerometry data shows the cumulative probabilities for acceleration of 1, 2, 3 milligravities, etc.   
This preprocess transformed them into probability mass on each acceleration value.  

```{r}
# Select users with compliance and generate accelerometry distribution

x <- readRDS("data/ukbiobank.rds")
mt <- read.csv("data/variable_table.csv")

# accelerometry variables
accel_vars <- names(x)[grep("fraction_acceleration", names(x), ignore.case = TRUE)]

# variables determining compliance
compliance_vars <- c("data_quality_good_wear_time_f90015_0_0", "data_quality_good_calibration_f90016_0_0", "data_quality_calibrated_on_own_data_f90017_0_0")

# We have a bunch of NAs.
xa <- x[,c("eid", accel_vars, compliance_vars)] |> na.omit() %>%
  filter(data_quality_good_wear_time_f90015_0_0 == "Yes", data_quality_good_calibration_f90016_0_0 == "Yes",
         data_quality_calibrated_on_own_data_f90017_0_0 == "Yes")

x_compliant <- x[x$eid %in% xa$eid, ]

ma <- as.matrix(select(xa, -eid, -compliance_vars))

for (j in seq_len(ncol(ma))[-1]) {
  ma[,j-1] <- ma[,j] - ma[,j-1]
}
ma <- ma[, -ncol(ma)]

accel_val <- map_dbl(accel_vars, ~ as.numeric(unlist(strsplit(.x, "_"))[3]))
accel_val <- accel_val[-length(accel_val)]

colnames(ma) <- factor(accel_val)
# ma is the matrix calculating percents of each accelerometry variable
```

```{r}
# Find max accelerometry value for each eid
tol = 1e-5
max_accel <- replicate(nrow(ma), 0)

for (i in seq_len(nrow(ma))) {
  j <- ncol(ma)
  stop <- FALSE
  while (!stop & j >= 1) {
    if (ma[i, j] > tol) {
      max_accel[i] <- accel_val[j]
      stop <- TRUE
    }
    j <- j-1
  }
}

# Find the number of modes in accelerometry distribution for each eid
# mode definition: the columns whose percentage is greater than its left column by threshhold = 0.005
find_mode <- function(x, thr = 0.005) {
  x <- c(0, x)
  res <- 0
  for (j in seq_len(length(x))[-1]) {
    if (x[j]-x[j-1] > thr) {
      res <- res + 1
    }
  }
  return (res)
} 

num_modes <- apply(ma, 1, find_mode)

# cutoffs: c(1, 20, 100)
# low_activity <- [1, 19], mid_activity <- [20, 95], high_activity <- [100, 1900]
```

```{r}
ma <- data.frame(ma)
ma <- ma %>% mutate(total_activity = rowSums(ma)) %>%
   mutate(low_activity = select(., X1:X19) %>% rowSums()) %>%
   mutate(mid_activity = select(., X20:X95) %>% rowSums()) %>%
   mutate(high_activity = select(., X100:X1900) %>% rowSums())
ma <- cbind(eid = factor(x_compliant$eid), ma)
```

### Probability Mass Table (for the first 500 participants)

```{r, layout="l-body-outset"}
tb1 <- ma[1:500, ] %>% select(eid:X1900)
paged_table(tb1)
```

# Accelerometry visualization (for the first 200 participants)

```{r}
create_trel_accel <- FALSE
if (create_trel_accel) {
accel_plt <- function(id) {
  accel_perc = as.numeric(ma[which(ma$eid == id), 
                             !(colnames(ma) %in% c("eid", "total_activity", "low_activity", "mid_activity", "high_activity"))])
  tibble(accel = factor(accel_val), accel_perc) |>
  ggplot(aes(x = accel, y = accel_perc)) +
  theme(axis.text.x = element_text(angle = 90, size = 5), text = element_text(size = 5)) +
  geom_col()
}

df <- tibble(eid = factor(ma$eid)[1:200], max_accel = factor(max_accel)[1:200], 
             num_modes = factor(num_modes)[1:200], plot = map_plot(eid, accel_plt))

df |> trelliscope(name = "Accelerometry Visualization", nrow = 1, ncol = 1, 
                   path = "assets/accelerometry-trelliscope")
} |> print()
knitr::include_app("assets/accelerometry-trelliscope/index.html")
```

### Interpretation

In the visualization, the probability mass functions follow a "3-peak" pattern.  
It has been validated for all users with compliance.  
3 peaks appear at 1, 20 and 200 milligravities.  

# Activity amount

Based on the acceleration cutoffs, the total probability within each range can measure the amount of different activities.  
1 - 19 milligravities: LOW intensity activity   
20 - 95 milligravities: MIDDLE intensity activity  
100 - 1900 milligravities: HIGH intensity activity

### Activity Amount Table

```{r, layout="l-body-outset"}
tb2 <- ma[1:500, ] %>% select(eid, total_activity:high_activity)
paged_table(tb2)
```
# Bivariate
## Accelerometry ~ Activity
### Categorical activity variables

```{r}
activity_baseline <- mt[mt$category == "Physical Activity" & mt$instance == 0, c(1,2,4)]
cat_act <- activity_baseline[activity_baseline$fieldid %in% 
           c(864, 884, 904, 843, 971, 991, 1011, 2624, 3637, 6164, 22032, 22035, 22036, 110005) ,1]

cat_act
```

### Activity amount vs. activity's categories
```{r}
reorder_levels <- c("Once in the last 4 weeks", "2-3 times in the last 4 weeks", 
                    "Once a week", "2-3 times a week", "4-5 times a week",
                    "Every day", "Do not know", "Prefer not to answer")

x_compliant$frequency_of_heavy_diy_in_last_4_weeks_f2624_0_0 <- 
  x_compliant$frequency_of_heavy_diy_in_last_4_weeks_f2624_0_0 %>% 
  factor(levels = reorder_levels)

x_compliant$frequency_of_light_diy_in_last_4_weeks_f1011_0_0 <-
  x_compliant$frequency_of_light_diy_in_last_4_weeks_f1011_0_0 %>%
  factor(levels = reorder_levels)

x_compliant$frequency_of_other_exercises_in_last_4_weeks_f3637_0_0 <-
  x_compliant$frequency_of_other_exercises_in_last_4_weeks_f3637_0_0 %>%
  factor(levels = reorder_levels)

x_compliant$frequency_of_strenuous_sports_in_last_4_weeks_f991_0_0 <-
  x_compliant$frequency_of_strenuous_sports_in_last_4_weeks_f991_0_0 %>%
  factor(levels = reorder_levels)

x_compliant$frequency_of_walking_for_pleasure_in_last_4_weeks_f971_0_0 <-
  x_compliant$frequency_of_walking_for_pleasure_in_last_4_weeks_f971_0_0 %>%
  factor(levels = reorder_levels)

```

```{r}
getPalette = colorRampPalette(brewer.pal(9, "Set1"))

create_act_cat_amount_trel <- FALSE
if (create_act_cat_amount_trel) {
act_amount_plot <- function(act) {
  plot_func <- function(act_amount) {
    df <- data.frame(ma[[act_amount]], factor(x_compliant[[act]])) %>% 
          na.omit %>% tibble
    colnames(df) <- c("act_amount", "activity")
    colourCount <- length(unique(df$activity))
    
    ggplot(df, aes(color=activity, y=act_amount, x=activity)) +
    geom_boxplot() + 
    xlab(act_amount) +
    scale_fill_manual(values = getPalette(colourCount)) + 
    theme(legend.position = "none", axis.ticks.x=element_blank(),
          axis.text.x=element_blank())
  }
  p1 <- plot_func("total_activity") +
       theme(legend.title = element_blank(), legend.text=element_text(size=5), 
             legend.key.size = unit(0.8, "cm"), legend.position = "top", 
             legend.box = "horizontal", legend.box.just = "center")
  p2 <- plot_func("low_activity")
  p3 <- plot_func("mid_activity")
  p4 <- plot_func("high_activity")
  p <- ggarrange(p1, p2, p3, p4, nrow = 2, ncol = 2, common.legend = TRUE, legend="bottom")
}

act_amount = tibble(var = cat_act, data_plot = map_plot(var, act_amount_plot))
act_amount %>% trelliscope(name = "Activity categories vs activity amount", nrow = 1, ncol = 1, 
                           path = "assets/act_cat_amount-trelliscope")
}
knitr::include_app("assets/act_cat_amount-trelliscope/index.html")
```

## Accelerometry ~ Sleep
### Variables related to sleep conditions

```{r}
med <- mt[[1]][mt$category == "Sleep" & mt$instance == 0] |>
  c(mt[[1]][mt$variable == "overall_health_rating_f2178_0_0"]) |>
  c(mt[[1]][mt$variable == "neuroticism_score_f20127_0_0"])
med
```

### Activity amount vs. sleep conditions

```{r}
create_med_act_trel <- FALSE
if (create_med_act_trel) {
med_act_plot <- function(med_var) {
  plot_func <- function(activity) {
    df <- data.frame(ma[[activity]], factor(x_compliant[[med_var]])) %>% 
          na.omit %>% tibble
    colnames(df) <- c("act_amount", "med_group")
    colourCount <- length(unique(df$med_group))
    
    ggplot(df, aes(color=med_group, y=act_amount, x=med_group)) +
    geom_boxplot() + 
    xlab(activity) +
    scale_fill_manual(values = getPalette(colourCount)) + 
    theme(legend.position = "none", axis.ticks.x=element_blank(),
          axis.text.x=element_blank())
  }
  p1 <- plot_func("total_activity") +
        theme(legend.title = element_blank(), legend.text=element_text(size=5), 
             legend.key.size = unit(0.8, "cm"), legend.position = "top", 
             legend.box = "horizontal", legend.box.just = "center")
  p2 <- plot_func("low_activity")
  p3 <- plot_func("mid_activity")
  p4 <- plot_func("high_activity")
  p <- ggarrange(p1, p2, p3, p4, nrow = 2, ncol = 2, common.legend = TRUE, legend="bottom")
}

med_act = tibble(var = med, data_plot = map_plot(var, med_act_plot))
med_act %>% trelliscope(name = "activity amount vs sleep conditions", nrow = 1, ncol = 1,
                        path = "assets/sleep_act-trelliscope")
}
knitr::include_app("assets/sleep_act-trelliscope/index.html")
```

### Statistical Analysis  

```{r}
activity <- rep(c("high_activity", "mid_activity", "low_activity"), 15)
med_var <- rep(c("daytime_dozing_sleeping_narcolepsy_f1220_0_0", "degree_bothered_by_trouble_sleeping_in_the_last_3_months_f21061_0_0", "did_your_sleep_change_f20532_0_0", "getting_up_in_morning_f1170_0_0", "morningevening_person_chronotype_f1180_0_0", "nap_during_day_f1190_0_0", "neuroticism_score_f20127_0_0", "overall_health_rating_f2178_0_0", "sleep_duration_f1160_0_0", "sleeping_too_much_f20534_0_0", "sleeplessness_insomnia_f1200_0_0", "snoring_f1210_0_0", "trouble_falling_asleep_f20533_0_0", "trouble_falling_or_staying_asleep_or_sleeping_too_much_f20517_0_0", "waking_too_early_f20535_0_0"), each=3)

best <- c("Never/rarely", "Not bothered at all", "No", "Very easy", "Definitely a 'morning' person", "Never/rarely", 0, "Excellent", 18, "No", "Never/rarely", "No", "No", "Not at all", "No")
best <- rep(best, each = 3)

worst <- c("All of the time", "Bothered a lot", "Yes", "Not at all easy", "Definitely an 'evening' person", "Usually", "11", "Poor", "-3", "Yes", "Usually", "Yes", "Yes", "Nearly every day", "Yes")
worst <- rep(worst, each = 3)

best_worst <- data.frame(activity, med_var, best, worst)

means <- rep(0, 45)
sds <- rep(0, 45)
pval <- rep(0, 45)

p_func <- function(activity, med_var, s1, s2) {
  df <- data.frame(ma[[activity]], factor(x_compliant[[med_var]])) %>% na.omit
  colnames(df) <- c("activity", "med_var")
  df <- df %>% tibble() %>%
    group_by(med_var) %>%
    group_by(med_var) %>%
    summarize(mean = mean(activity),
              sd = sd(activity)/sqrt(length(activity)))
  
  xdiff = df$mean[df$med_var == s1] - df$mean[df$med_var == s2]
  xsd = sqrt(df$sd[df$med_var == s1]^2 + df$sd[df$med_var == s2]^2)
  pt = tibble(x = seq(-.1, .1, 0.001), 
              y = dnorm(x, mean = xdiff, sd = xsd))
  pval <- min(
    round(pnorm(0, mean = xdiff, sd = xsd, lower.tail = FALSE), 4),
    round(pnorm(0, mean = xdiff, sd = xsd), 4)
  )
  return (c(xdiff, xsd, pval))
}

for (i in 1:45) {
  activity <- best_worst[i, 1]
  med_var <- best_worst[i, 2]
  s1 <- best_worst[i, 3]
  s2 <- best_worst[i, 4]
  res <- p_func(activity, med_var, s1, s2)
  means[i] <- res[1]
  sds[i] <- res[2]
  pval[i] <- res[3]
}

best_worst <- cbind(best_worst, means) %>% cbind(sds) %>% cbind(pval)
```

According to Central Limit Theorem, mean of users' activity amount with medical condition ('best' or 'worst') approximates a normal distribution.  
Thus the mean of activity amount difference between 'best' and 'worst' should also has a normal distribution.  
p-value = minimal { p-value(mean difference > 0), p-value(mean difference < 0) }     
If p-value << 0.5, the activity amount difference between 'best' and 'worst' is significant.  

```{r}
create_normal_trel <- FALSE
if (create_normal_trel) {
  normal_plot_func <- function(activity, med_var) {
    i <- which(best_worst[, 1] == activity & best_worst[, 2] == med_var)
    pt = tibble(x = seq(-.1, .1, 0.001), 
                y = dnorm(x, mean = best_worst[i, "means"], sd = best_worst[i, "sds"]))
    ggplot(pt, aes(x = x, y = y)) + geom_line() + xlab(best_worst[i, "med_var"]) +
    geom_vline(xintercept = 0, linetype="dotted", color = "salmon", size=1) +
    xlab("mean of ('best' - 'worst')") +
    ylab("probability density") +
    ggtitle(paste0("p-value = ", best_worst[i, "pval"])) +
    theme(plot.title = element_text(hjust = 0.5))
 }
  best_worst <- best_worst %>% mutate(normal_plot = map2_plot(activity, med_var, normal_plot_func))
  best_worst %>% trelliscope(name = "Normal distribution_mean_best_worst_diff", nrow = 1, ncol = 1,
                             path = "assets/med_normal-trelliscope")
}
knitr::include_app("assets/med_normal-trelliscope/index.html")
```

