---
title: "Prediction"
description: |
  Incorporating accelerometry data to predict sleeping conditions
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(randomForest)
```

```{r}
sleep_profile <- readRDS("data/sleep_profile.rds")
df <- readRDS("data/ukbiobank.rds")
```

```{r}
sleep_profile <- sleep_profile[sleep_profile$louvain %in% c(4, 6, 7, 3, 5, 2, 11, 8, 9, 1, 16), ]
```

```{r}
var_name <- c("eid", "weight_f21002_0_0", "standing_height_f50_0_0", 
              "year_of_birth_f34_0_0", "sex_f31_0_0")
x_df <- df[, var_name] |>
  na.omit() 

x_df <- x_df[x_df$eid %in% sleep_profile$eid, ]
```

```{r}
# Define a function to perform z-score standardization
z_score_standardization <- function(x) {
  return((x - mean(x)) / sd(x))
}

# Apply z-score standardization to specific columns in the data frame
x_df$weight_f21002_0_0 <- z_score_standardization(x_df$weight_f21002_0_0)
x_df$standing_height_f50_0_0 <- z_score_standardization(x_df$standing_height_f50_0_0)
x_df$year_of_birth_f34_0_0 <- z_score_standardization(x_df$year_of_birth_f34_0_0)
```

```{r}
new_df <- merge(sleep_profile, x_df, by = "eid", all = FALSE)
new_df$louvain <- as.factor(new_df$louvain)
new_df <- new_df |> na.omit()
```

```{r}
set.seed(123)  # Set seed for reproducibility
train_indices <- sample(1:nrow(x_df), nrow(new_df) * 0.8)  # Randomly select 80% of indices
train_data <- new_df[train_indices, ]  # Subset train data
test_data <- new_df[-train_indices, ]  # Subset test data
```

## Predict sleep outcome using only demographic variables
```{r}
# Specify the predictor variables and the target variable
predictors <- c("weight_f21002_0_0", "standing_height_f50_0_0", 
              "year_of_birth_f34_0_0", "sex_f31_0_0")
target <- "louvain"

# Train the Random Forest model
rf_model1 <- randomForest(formula = as.formula(paste(target, "~", paste(predictors, collapse = " + "))),
                         data = train_data, ntree = 500, mtry = sqrt(length(predictors)))

# Print the summary of the trained model
print(rf_model1)
```

```{r}
# Make predictions on the test data
predictions <- predict(rf_model1, newdata = test_data)

# Evaluate the model performance
confusion_matrix <- table(predictions, test_data$louvain)
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(confusion_matrix)
print(paste("Accuracy:", round(accuracy, 2)))
```

```{r}
accel_profile <- readRDS("data/accel_profile.rds")
```

```{r}
a_df <- merge(new_df, accel_profile, by = "eid", all = FALSE)
a_df$louvain <- as.factor(a_df$louvain)
a_df <- a_df |> na.omit()
```

```{r}
a_df$low_activity <- z_score_standardization(a_df$low_activity)
a_df$mid_activity <- z_score_standardization(a_df$mid_activity)
a_df$high_activity <- z_score_standardization(a_df$high_activity)
```

```{r}
a_train_data <- a_df[train_indices, ]  # Subset train data
a_test_data <- a_df[-train_indices, ]  # Subset test data
```

## Predict sleep outcome using demographic variables + accelerometry
```{r}
# Specify the predictor variables and the target variable
a_predictors <- c("weight_f21002_0_0", "standing_height_f50_0_0", 
              "year_of_birth_f34_0_0", "sex_f31_0_0", "low_activity", 
              "mid_activity", "high_activity")

# Train the Random Forest model
rf_model2 <- randomForest(formula = as.formula(paste(target, "~", paste(a_predictors, collapse = " + "))),
                         data = a_train_data, ntree = 500, mtry = sqrt(length(predictors)))

# Print the summary of the trained model
print(rf_model2)
```

```{r}
# Make predictions on the test data
a_predictions <- predict(rf_model2, newdata = a_test_data)

# Evaluate the model performance
confusion_matrix <- table(a_predictions, a_test_data$louvain)
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(confusion_matrix)
print(paste("Accuracy:", round(accuracy, 2)))
```

### Predictor significance
```{r}
var_importance <- importance(rf_model2)
var_importance_sorted <- var_importance[order(var_importance, decreasing = TRUE),]
print(var_importance_sorted)
```

Incorporating activity scores calculated from accelerometry data can increase accuracy rate of our random forest model by 18.75%. Activity score variables turn out to show significance compared to other demographic variables. Based on our findings, we can infer that accelerometry data has the potential to serve as an additional tool for predicting sleep conditions.

However, the accuracy rate is still low. This is because the independent variables (especially the activity scores) represent summary data, which may have resulted in the loss of significant information during the aggregation process." This can be solved in the future by taking full advantage of raw accelerometry data (beyond our current scope).